# Population structure and relatedness

Population structure due to genetic ancestry and genetic correlation due to sample relatedness are important factors to consider when performing association tests as their presence can lead to mis-calibration of test statistics and false positive associations. This tutorial demonstrates how to perform population structure and relatedness inference using the [GENESIS](https://bioconductor.org/packages/release/bioc/html/GENESIS.html) and [SNPRelate](https://www.bioconductor.org/packages/release/bioc/html/SNPRelate.html) R/Bioconductor packages. We use the results of this inference to perform association tests in the mixed model tutorial.

## Computing a GRM

We can use the [SNPRelate](https://github.com/zhengxwen/SNPRelate) package to compute a Genetic Relationship matrix (GRM) using GDS files.  A GRM captures genetic correlation among samples due to both distant ancestry (i.e. population structure) and recent kinship (i.e. familial relatedness) in a single matrix.

SNPRelate offers several algorithms for computing a GRM, including the commonly-used GCTA [Yang et al 2011](https://www.ncbi.nlm.nih.gov/pubmed/21167468) (`method = "GCTA"`) and allelic matching based estimators described by [Weir and Goudet 2017](https://www.ncbi.nlm.nih.gov/pubmed/28550018) (`method = "IndivBeta"`).

```{r grm}
library(SeqArray)
repo_path <- "https://github.com/smgogarten/GENESIS_workshop_2026/raw/main"
if (!dir.exists("data")) dir.create("data")
gdsfile <- "data/1KG_phase3_subset_chr1.gds"
if (!file.exists(gdsfile)) download.file(file.path(repo_path, gdsfile), gdsfile)
gdsfmt::showfile.gds(closeall=TRUE) # make sure file is not already open
gds <- seqOpen(gdsfile)

library(SNPRelate)
grm <- snpgdsGRM(gds, method="GCTA")
names(grm)
dim(grm$grm)

# look at the top corner of the matrix
grm$grm[1:5,1:5]

seqClose(gds)
```


## PC-Relate

To disentangle distant ancestry (i.e. population structure) from recent kinship (i.e. familial relatedness), we implement the analysis described in [Conomos et al., 2016](https://www.cell.com/ajhg/fulltext/S0002-9297(15)00496-6). This approach uses the [KING-robust](http://www.ncbi.nlm.nih.gov/pubmed/20926424), [PC-AiR](http://www.ncbi.nlm.nih.gov/pubmed/25810074), and [PC-Relate](http://www.ncbi.nlm.nih.gov/pubmed/26748516) methods.

### LD-pruning

We generally advise that population structure and relatedness inference be performed using a set of (nearly) independent genetic variants. To find this set of variants, we perform linkage-disequilibrium (LD) pruning on the study sample set. We typically use an LD threshold of `r^2 < 0.1` to select variants. We use the [SNPRelate](https://github.com/zhengxwen/SNPRelate) package to perform LD-pruning with GDS files. 

```{r ld-pruning}
# use a GDS file with all chromosomes
library(SeqArray)
gdsfile <- "data/1KG_phase3_subset.gds"
if (!file.exists(gdsfile)) download.file(file.path(repo_path, gdsfile), gdsfile)
gdsfmt::showfile.gds(closeall=TRUE) # make sure file is not already open
gds <- seqOpen(gdsfile)

# use a subset of 100 samples to make things run faster
sampfile <- "data/samples_subset100.rds"
if (!file.exists(sampfile)) download.file(file.path(repo_path, sampfile), sampfile)
sample.id <- readRDS(sampfile)

# LD pruning to get variant set
library(SNPRelate)
set.seed(100) # LD pruning has a random element; so make this reproducible
snpset <- snpgdsLDpruning(gds, sample.id=sample.id, method="corr", 
                          slide.max.bp=10e6, ld.threshold=sqrt(0.1))
sapply(snpset, length)
pruned <- unlist(snpset, use.names=FALSE)
```

### KING-robust

Step 1 is to get initial kinship estimates using [KING-robust](http://www.ncbi.nlm.nih.gov/pubmed/20926424), which is robust to discrete population structure but not ancestry admixture. KING-robust will be able to identify close relatives (e.g. 1st and 2nd degree) reliably, but may identify spurious pairs or miss more distant pairs of relatives in the presence of admixture. KING is available as its own software, but the KING-robust algorithm is also available in SNPRelate via the function `snpgdsIBDKING`.

```{r king}
# KING
king <- snpgdsIBDKING(gds, sample.id=sample.id, snp.id=pruned)
names(king)

# extract the kinship estimates
kingMat <- king$kinship
colnames(kingMat) <- rownames(kingMat) <- king$sample.id
dim(kingMat)
# look at the top corner of the matrix
kingMat[1:5,1:5]
```

We extract pairwise kinship estimates and IBS0 values (the proportion of variants for which the pair of indivdiuals share 0 alleles identical by state) to plot.

```{r king_plot}
kinship <- snpgdsIBDSelection(king)
head(kinship)

library(ggplot2)
ggplot(kinship, aes(IBS0, kinship)) +
    geom_hline(yintercept=2^(-seq(3,9,2)/2), linetype="dashed", color="grey") +
    geom_point(alpha=0.5) +
    ylab("kinship estimate") +
    theme_bw()
```

We see a few parent-offspring, full sibling, 2nd degree, and 3rd degree relative pairs. The abundance of negative estimates represent pairs of individuals who have ancestry from different populations -- the magnitude of the negative relationship is informative of how different their ancestries are; more on this below.

### PC-AiR

The next step is [PC-AiR](http://www.ncbi.nlm.nih.gov/pubmed/25810074), which provides robust population structure inference in samples with kinship and pedigree structure. PC-AiR is available in the GENESIS package via the function `pcair`. The steps in `pcair` can also be performed individually, which is illustrated below.

First, PC-AiR partitions the full sample set into a set of mutually unrelated samples that is maximally informative about all ancestries in the sample (i.e. the unrelated set) and their relatives (i.e. the related set). We use a 3rd degree kinship threshold (`kin.thresh = 2^(-9/2)`), which corresponds to first cousins -- this defines anyone less related than first cousins as "unrelated". We use the negative KING-robust estimates as "ancestry divergence" measures (`divMat`) to identify pairs of samples with different ancestry -- we preferentially select individuals with many negative estimates for the unrelated set to ensure ancestry representation. For now, we also use the KING-robust estimates as our kinship measures (`kinMat`); more on this below.

Once the unrelated and related sets are identified, PC-AiR performs a standard Principal Component Analysis (PCA) on the unrelated set of individuals and then projects the relatives onto the PCs. Under the hood, PC-AiR uses the SNPRelate package for efficient PC computation and projection.

```{r pcair_partition}
library(GENESIS)
sampset <- pcairPartition(kinobj=kingMat, kin.thresh=2^(-9/2),
                          divobj=kingMat, div.thresh=-2^(-9/2))
names(sampset)
sapply(sampset, length)
```

Using the SNPRelate package, we run PCA on the unrelated set and project values for the related set. We use the same LD pruned set of variants again.

```{r pcair_1}
# run PCA on unrelated set
pca.unrel <- snpgdsPCA(gds, sample.id=sampset$unrels, snp.id=pruned)

# project values for relatives
snp.load <- snpgdsPCASNPLoading(pca.unrel, gdsobj=gds)
samp.load <- snpgdsPCASampLoading(snp.load, gdsobj=gds, sample.id=sampset$rels)

# combine unrelated and related PCs and order as in GDS file
pcs <- rbind(pca.unrel$eigenvect, samp.load$eigenvect)
rownames(pcs) <- c(pca.unrel$sample.id, samp.load$sample.id)
samp.ord <- match(sample.id, rownames(pcs))
pcs <- pcs[samp.ord,]
```

We need to determine which PCs are ancestry informative. To do this we need population information for the 1000 Genomes samples. This information is stored in an `AnnotatedDataFrame`, which is a data.frame with optional metadata describing the columns. The class is defined in the Biobase package. We load the stored object.

```{r annot}
library(Biobase)
sampfile <- "data/sample_annotation.rds"
if (!file.exists(sampfile)) download.file(file.path(repo_path, sampfile), sampfile)
annot <- readRDS(sampfile)
annot
head(pData(annot))
varMetadata(annot)
```

We make a parallel coordinates plot, color-coding by 1000 Genomes population. We load the [dplyr](http://dplyr.tidyverse.org) package for data.frame manipulation.

```{r pcair_parcoord}
pc.df <- as.data.frame(pcs)
names(pc.df) <- 1:ncol(pcs)
pc.df$sample.id <- row.names(pcs)

library(dplyr)
annot <- pData(annot) %>%
        dplyr::select(sample.id, Population)
pc.df <- left_join(pc.df, annot, by="sample.id")

library(GGally)
library(RColorBrewer)
pop.cols <- setNames(brewer.pal(12, "Paired"),
                 c("ACB", "ASW", "CEU", "GBR", "CHB", "JPT", "CLM", "MXL", "LWK", "YRI", "GIH", "PUR"))
ggparcoord(pc.df, columns=1:12, groupColumn="Population", scale="uniminmax") +
    scale_color_manual(values=pop.cols) +
    xlab("PC") + ylab("")
```


### PC-Relate

The first 2 PCs separate populations, so we use them to compute kinship estimates adjusting for ancestry. The PC-Relate function expects a `SeqVarIterator` object.

```{r pcrelate_1}
seqResetFilter(gds, verbose=FALSE)
library(SeqVarTools)
seqData <- SeqVarData(gds)
seqSetFilter(seqData, variant.id=pruned)
iterator <- SeqVarBlockIterator(seqData, verbose=FALSE)

pcrel <- pcrelate(iterator, pcs=pcs[,1:2], training.set=sampset$unrels, 
                  sample.include=sample.id)
names(pcrel)
```

PC-Relate is an iterative method. Now that we have ancestry-adjusted kinship estimates, we can use them to better adjust for ancestry in the PCs. This time we use the `pcair` function, which combines partitioning the sample set and running PCA in one step. First we need to make a kinship matrix from the PC-Relate results. The KING matrix is still used for ancestry divergence.

```{r pcair_2}
pcrelMat <- pcrelateToMatrix(pcrel, scaleKin=1, verbose=FALSE)

seqResetFilter(seqData, verbose=FALSE)
pca <- pcair(seqData,
             kinobj=pcrelMat, kin.thresh=2^(-9/2),
             divobj=kingMat, div.thresh=-2^(-9/2),
             sample.include=sample.id, snp.include=pruned,
             verbose=FALSE)
names(pca)

pcs <- pca$vectors
pc.df <- as.data.frame(pcs)
names(pc.df) <- paste0("PC", 1:ncol(pcs))
pc.df$sample.id <- row.names(pcs)
pc.df <- left_join(pc.df, annot, by="sample.id")

ggplot(pc.df, aes(PC1, PC2, color=Population)) + geom_point() +
    scale_color_manual(values=pop.cols)
```

Now we use the revised PCs to compute new kinship estimates. One can run the iteration multiple times and check for conversion, but usually two rounds are sufficient.

```{r pcrelate_2}
seqSetFilter(seqData, variant.id=pruned)
iterator <- SeqVarBlockIterator(seqData, verbose=FALSE)
pcrel <- pcrelate(iterator, pcs=pcs[,1:2], training.set=pca$unrels, 
                  sample.include=sample.id)
saveRDS(pcrel, file="data/pcrelate_kinship.rds")
```

We plot the kinship estimates from PC-Relate, and notice that the values for less related pairs are much better behaved.

```{r pcrelate_plot}
kinship <- pcrel$kinBtwn

ggplot(kinship, aes(k0, kin)) +
    geom_hline(yintercept=2^(-seq(3,9,2)/2), linetype="dashed", color="grey") +
    geom_point(alpha=0.5) +
    ylab("kinship estimate") +
    theme_bw()
```

```{r close}
seqClose(gds)
```


### Comparison with pedigree

We can detect pedigree errors and sample identity problems by comparing the pedigree with empirical kinship estimates. We use a function from the GWASTools package, `pedigreePairwiseRelatedness`, to get expected pairwise relationships based on the pedigree.

```{r pedigree}
pedfile <- "data/pedigree.rds"
if (!file.exists(pedfile)) download.file(file.path(repo_path, pedfile), pedfile)
ped <- readRDS(pedfile)
head(ped)
pw <- GWASTools::pedigreePairwiseRelatedness(ped)
names(pw)
rel <- pw$relativeprs
head(rel)
table(rel$relation)
distinct(rel, relation, kinship) %>%
    arrange(-kinship)

## assign degrees to expected relationship pairs
rel <- rel %>%
    mutate(exp.rel=ifelse(kinship == 0.125, "Deg2", ifelse(kinship == 0.0625, "Deg3", relation)),
           pair=GWASTools::pasteSorted(Individ1, Individ2)) %>%
    select(pair, family, relation, exp.rel)

## assign degrees to observed relationship pairs
cut.dup <- 1/(2^(3/2))
cut.deg1 <- 1/(2^(5/2))
cut.deg2 <- 1/(2^(7/2))
cut.deg3 <- 1/(2^(9/2))
cut.k0 <- 0.1
kinship <- kinship %>%
    mutate(obs.rel=ifelse(kin > cut.dup, "Dup",
                   ifelse(kin > cut.deg1 & k0 < cut.k0, "PO",
                   ifelse(kin > cut.deg1, "FS",
                   ifelse(kin > cut.deg2, "Deg2",
                   ifelse(kin > cut.deg3, "Deg3", "U"))))))
table(kinship$obs.rel)

# merge observed and expected relationships
kin.obs <- kinship %>%
    select(ID1, ID2, kin, k0, obs.rel) %>%
    mutate(pair=GWASTools::pasteSorted(ID1, ID2)) %>%
    left_join(rel, by="pair") %>%
    select(-pair) %>%
    mutate(exp.rel=ifelse(is.na(exp.rel), "U", exp.rel)) %>%
    filter(!(exp.rel == "U" & obs.rel == "U"))
table(kin.obs$exp.rel, kin.obs$obs.rel)

ggplot(kin.obs, aes(k0, kin, color=obs.rel)) + geom_point()
```

All the observed relationships were unexpected. These samples are from 1000 Genomes sequencing, and known relatives were excluded from the released data. Here we have detected some cryptic relatives that were not annotated in the pedigree.

### Advanced Notes

In large samples (tens of thousands or more), the KING-robust and PC-Relate analyses can be very time consuming, so we suggest the following alternative procedure for deconvoluting ancestry and relatedness:

1. Use the [KING-IBDseg method](https://www.kingrelatedness.com/manual.shtml#IBDSEG) to estimate kinship for all pairs of individuals. This method uses a fast algorithm to approximate IBD segments based on identity by state and works well in the presence of ancestry admixture -- in our experience, it gives very similar results to PC-Relate. The software to run KING-IBDseg uses PLINK files.
2. Perform a PC-AiR analysis to infer population structure using the KING-IBDseg estimates as the kinship estimates and using no ancestry divergence measures (in large samples, the PC-AiR algorithm works well even without the ancestry divergence).

## Exercise

Complete one round of iteration using all samples from the test dataset and plot the results. Be sure to examine the parallel coordinates plot to determine the appropriate number of PCs to give as an argument to `pcrelate`.
